---
title: "Loan_Defaults"
author: "ChrisCarney_LisaStimpson_JaeChoi_YiHan_NiklasRikala"
date: "4/7/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
INTRO
  The objective of this analysis is to utilize data sourced from a peer-to-peer lending service to reliably determine when a borrower will default on a loan. In order to achieve this, the data must be examined, subjectivley and objectivley, with an understanding of loans and basic finance to determine which factors will and will not be valuable to predicting loan defults. The data consists of 52 variables regarding information about the borrowers current and past financial, credit, residential, and employment history.First we will examine and clean the data. 

```{r}
loans<-read.csv('Listings2013.csv')
str(loans)
summary(loans)
#Duplicate Information
  #loan_status-desription - Remove status because it hold less information
loans$loan_status<-NULL
  #income_range-description - Remove Range because it holds less information
loans$income_range<-NULL
#Non-sensical Values
  #DTI--- 1000000 --- investigate
dti_invest<-subset(loans,loans$dti_wprosper_loan==1000000)#DTI is set to 1000000 when Income Verifiable is False -- How do we handle this? Subsitute Stated income to calculate this?
  #MonthsEmployed --- negative values?
months_emp_invest<-subset(loans,loans$months_employed<0) #Doesn't make sence how should we handle this? Take absolute value?
loans$months_employed<-abs(loans$months_employed)
summary(loans$months_employed)
  #bankcard Utilization -- Can this value be over 1?
  
#Missing Values
  #Installment Values
installment_invest<-subset(loans,is.na(loans$installment_balance)) #How should we handle? Set to 0 if installments not available?
months_NA_inv<-subset(loans,is.na(loans$months_employed)) #How do we handle these? Occupation is mostly Null Here
loans<-na.omit(loans) #Removed for Now

#Reformatting
#Dates
library(lubridate)
loans$loan_origination_date<-mdy(loans$loan_origination_date)
loans$first_recorded_credit_line<-mdy(loans$first_recorded_credit_line)
#Factors
loans$prosper_score<-factor(loans$prosper_score)
loans$listing_category_id<-factor(loans$listing_category_id)
loans$lender_indicator<-factor(loans$lender_indicator)
#Remove Useless Data
loans$borrower_city<-NULL
loans$borrower_state<-NULL

#Check for NA and NULL
apply(loans, 2, function(x) any(is.na(x) | is.null(x)))

#Create Default variable
loans$default<-ifelse(loans$loan_status_description=='DEFAULTED',1,0)
```
DATA CLEANING
  After loading the data set and examining its structure and a summary of each variable some discrepancies were found. There was duplicate varaibles for the loan status and income range values, the versions that contained more complete information were then selected and the other verisons removed. There were non sensical values that needed to be investigated including values of 1000000 in DTI ratios (found to mean income not verifiable or income is 0 and therefore the ratio goes to infinity) and negative Months of employment for people listed as employed (no significant understanding found). Thus far, it has been decided that 1000000 is a reasonable representation of infinity and therefore it will remain (for now). To handle negative months of employment, the absolute value of these numbers was taken to remove thier negative effect (IDK if this is the right approach). I had a question about bankcard utilization and how it is calculated... Can bankcard utiliation be greater than 1? To get a better sense of this datas significance, a linear regression will be utilized.

```{r}
#Linear Regression
library(lmtest)
loans.m1<-lm(formula = default ~ . , data = loans)
summary(loans.m1) #Need to so some removing of conflicting varaibles
```




